---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    cache = TRUE,
    warning = FALSE,
    message = FALSE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
    )
```

# Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models

<!-- badges: start -->
<!-- badges: end -->

The goal of `neurogam` is to provide utilities for estimating onset and offset of effects in M/EEG data.

## Installation

You can install the development version of `neurogam` from GitHub with:

```{r, eval = FALSE}
install.packages("remotes")

remotes::install_github(
    repo = "https://github.com/lnalborczyk/neurogam",
    dependencies = TRUE
    )
```

## Usage

Below we fit a Bayesian generalised additive multilevel model (BGAMM) to estimate the onset and offset of effect from simulated EEG data.

```{r, results = "hide"}
# loading the neurogam package
library(neurogam)

# importing some simulated EEG data
data(eeg_data)
head(eeg_data)

# fitting the BGAMM to identify clusters
results <- testing_through_time(data = eeg_data, post_prob_ratio_threshold = 3)
```

```{r}
# displaying the identified clusters
print(results$clusters)
```
