---
title: "Application to 1D temporal EEG data"
author: "Ladislas Nalborczyk"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Application to 1D temporal EEG data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    cache = TRUE,
    collapse = TRUE,
    warning = FALSE,
    message = FALSE,
    comment = "#>",
    out.width = "100%",
    dev = "png",
    dpi = 300
    )
```

## Importing and visualising EEG data

Below we import and reshape EEG data from the [`eegkit` package](https://cran.r-project.org/web/packages/eegkit/index.html).

```{r data-import, fig.width = 12, fig.height = 12}
library(neurogam)
library(ggplot2)
library(eegkit)
library(dplyr)

# retrieving some EEG data
data(eegdata)
head(eegdata)

# plotting the average ERP per group and channel
eegdata |>
    summarise(voltage = mean(voltage), .by = c(group, channel, time) ) |>
    ggplot(aes(x = time, y = voltage, colour = group) ) +
    geom_line() +
    facet_wrap(~channel) +
    theme_bw()
```

```{r data-reshape}
# reshape the data
eeg_data <- eegdata |>
    # keeping only one channel
    dplyr::filter(channel == "PZ") |>
    # converting timesteps to seconds
    mutate(time = (time + 1) / 256) |>
    # rounding numeric variables
    mutate(across(where(is.numeric), ~ round(.x, 4) ) ) |>
    # removing NAs
    na.omit()

# show a few rows
head(eeg_data)
```

## Fitting the model

We fit the model (a BGAMM) on one channel (PZ) to test for group difference at every timestep.

```{r clusters, results = "hide"}
# fitting the BGAMM to identify clusters (around 30 min on 4 parallel apple M4 cores)
results <- testing_through_time(
    # EEG data
    data = eeg_data,
    # participant column
    participant_id = "subject",
    # EEG column
    outcome_id = "voltage",
    # name of predictor in data
    predictor_id = "group",
    # basis dimension
    kvalue = 40,
    # we recommend fitting the GAMM with summary statistics (mean and SD)
    multilevel = "summary",
    # threshold on posterior odds
    threshold = 10,
    # number of MCMCs
    chains = 4,
    # number of parallel cores
    cores = 4
    )
```

### Visualising the results

```{r print-clusters}
# displaying the identified clusters
print(results)
```

```{r clusters-plot, fig.width = 9, fig.height = 6}
# plotting the data, model's predictions, and clusters
plot(results)
```

### Computing clusters at the participant-level

Below we fit a new model, specifying `predictor_id = NA` (because `group` varies across participants) and `by_ppt = TRUE` to test whether EEG amplitude (`voltage`) differs from 0 and return clusters at the participant level.

```{r clusters-ppt, results = "hide"}
# fitting the BGAMM to identify clusters (around 30 min on 4 parallel apple M4 cores)
results <- testing_through_time(
    # EEG data
    data = eeg_data,
    # participant column
    participant_id = "subject",
    # EEG column
    outcome_id = "voltage",
    # here we use no predictor (because group varies across participants)
    predictor_id = NA,
    # basis dimension (for both the group and participant levels)
    kvalue = 40,
    # we recommend fitting the GAMM with summary statistics (mean and SD)
    multilevel = "summary",
    # return clusters at both the group and participant levels
    by_ppt = TRUE,
    # threshold on posterior odds
    threshold = 10,
    # number of MCMCs
    chains = 4,
    # number of parallel cores
    cores = 4
    )
```

### Visualising the results

```{r print-clusters-ppt}
# displaying the identified clusters
print(results)
```

```{r clusters-ppt-plot, fig.width = 12, fig.height = 9}
# plotting the data, model's predictions, and clusters
plot(results)
```

### Posterior predictive checks

We recommend visually assessing the predictions of the model against the observed data (for each participant). We provide a lightweight `ppc()` method, but you can conduct various PPCs with `brms::pp_check(results$model, ...)` (for all available PPCs, see <https://mc-stan.org/bayesplot/reference/PPC-overview.html>).

```{r ppc1-plot, fig.width = 12, fig.height = 6}
# posterior predictive checks (PPCs) at the group level
ppc(object = results, ppc_type = "group")
```

```{r ppc2-plot, fig.width = 12, fig.height = 9}
# posterior predictive checks (PPCs) at the participant level
ppc(object = results, ppc_type = "participant")
```
