---
title: "Checking and modelling residual auto-correlation (1D temporal data)"
author: "Ladislas Nalborczyk"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Checking and modelling residual auto-correlation (1D temporal data)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    cache = TRUE,
    collapse = TRUE,
    warning = FALSE,
    message = FALSE,
    comment = "#>",
    out.width = "100%",
    dev = "png",
    dpi = 300
    )
```

## Importing and visualising EEG data

Below we import and reshape EEG data from the [`eegkit` package](https://cran.r-project.org/web/packages/eegkit/index.html).

```{r data-import, fig.width = 12, fig.height = 12}
library(patchwork)
library(neurogam)
library(ggplot2)
library(eegkit)
library(dplyr)

# retrieving some EEG data
data(eegdata)
head(eegdata)

# plotting the average ERP per group and channel
eegdata |>
    summarise(voltage = mean(voltage), .by = c(group, channel, time) ) |>
    ggplot(aes(x = time, y = voltage, colour = group) ) +
    geom_line() +
    facet_wrap(~channel) +
    theme_bw()
```

```{r data-reshape}
# reshape the data
eeg_data <- eegdata |>
    # keeping only one channel
    dplyr::filter(channel == "PZ") |>
    # converting timesteps to seconds
    mutate(time = (time + 1) / 256) |>
    # rounding numeric variables
    mutate(across(where(is.numeric), ~ round(.x, 4) ) ) |>
    # removing NAs
    na.omit()

# show a few rows
head(eeg_data)
```

## Fitting a first model

We fit the model (a BGAMM) on one channel (PZ) to test for group difference at every timestep.

```{r clusters1, results = "hide"}
# fitting the BGAMM to identify clusters
results1 <- testing_through_time(
    # EEG data
    data = eeg_data,
    # participant column
    participant_id = "subject",
    # EEG column
    outcome_id = "voltage",
    # no predictor
    predictor_id = NA,
    # basis dimension
    kvalue = 20,
    # we recommend fitting the GAMM on summary statistics (mean and SD)
    multilevel = "summary",
    # no by-participant varying smooth (for speed purposes)
    varying_smooth = FALSE,
    # threshold on posterior odds
    threshold = 10,
    # number of MCMCs
    chains = 4,
    # number of parallel cores
    cores = 4,
    # number of warmup iterations
    warmup = 1000,
    # total number of iterations (per chain)
    iter = 2000
    )
```

### Visualising the results

```{r print-clusters1}
# displaying the identified clusters
print(results1)
```

```{r clusters-plot1, fig.width = 9, fig.height = 6}
# plotting the data, model's predictions, and clusters
plot(results1)
```

### Checking auto-correlation

Below we use the `check_residual_autocorrelation()` function to check to residual auto-correlation. The ACF plot (per participant) reveals strong residual auto-correlation.

```{r check-autocor1, fig.width = 9, fig.height = 6}
# checking residual auto-correlation
ar_check1 <- check_residual_autocorrelation(
    fit = results1$model
    )

# residual ACF (for some participants)
ar_check1$plots$acf_raw
```

Compare with the residuals of the equivalent `mgcv` model (see also [this vignette](https://cran.r-project.org/web/packages/itsadug/vignettes/acf.html)).

```{r itsadug, fig.width = 9, fig.height = 6}
library(itsadug)
library(mgcv)

# fitting an equivalent mgcv model
m1 <- bam(voltage ~ s(time, k = 20) + s(subject, bs = "re"), data = eeg_data)

# plotting the raw residuals for some participants
acf_resid(model = m1, split_pred = "subject", n = 9)
```

## Modelling auto-correlation

The `testing_through_time()` function allows including a latent AR1 model to capture unmodelled autocorrelation (via `include_ar_term = TRUE`), which is estimated simultaneously to the other model parameters. This contrast with conventional approaches using a 2-step procedure consisting in first estimating the residual auto-correlation (from a first model), and then including this correlation as a fixed parameter in a subsequent model. Our approach, while more computationally intense, has several advantages:

* **Proper uncertainty propagation.** Estimating the latent AR term jointly with the smooth/linear effects propagates uncertainty about autocorrelation into *all* model parameters and predictions. In contrast, a 2-step "estimate then fix it" approach is a plug-in procedure that conditions on an uncertain quantity, which may yield **overconfident** intervals (too-small SEs / under-coverage). 

* **Less bias from mean-correlation confounding.** Residual autocorrelation depends on what the mean model has already explained (and on smoothing penalties). When estimating autocorrelation from residuals of an initial fit, we risk **misallocating structure** between the smooth trend and the AR process. Joint estimation helps the model decide, probabilistically, whether variation is better explained by the smooth component or by autocorrelated deviations.

* **Cleaner, less error-prone prediction.** Two-step workflows often require awkward post-hoc steps (fit trend, extract residuals, fit AR model, forecast residuals, back-transform, etc.), which is tedious and easy to get wrong. A joint latent AR formulation gives a single coherent generative model for estimation *and* prediction (see also [this blogpost](https://ecogambler.netlify.app/blog/autocorrelated-gams/)).

```{r clusters2, results = "hide"}
# fitting the BGAMM to identify clusters
results2 <- testing_through_time(
    # EEG data
    data = eeg_data,
    # participant column
    participant_id = "subject",
    # EEG column
    outcome_id = "voltage",
    # no predictor
    predictor_id = NA,
    # basis dimension
    kvalue = 20,
    # we recommend fitting the GAMM on summary statistics (mean and SD)
    multilevel = "summary",
    # no by-participant varying smooth (for speed purposes)
    varying_smooth = FALSE,
    # including (estimating) auto-correlation within the same model
    include_ar_term = TRUE,
    # threshold on posterior odds
    threshold = 10,
    # number of MCMCs
    chains = 4,
    # number of parallel cores
    cores = 4,
    # number of warmup iterations
    warmup = 1000,
    # total number of iterations (per chain)
    iter = 2000
    )
```

```{r check-autocor2, fig.width = 9, fig.height = 6}
# checking residual auto-correlation
ar_check2 <- check_residual_autocorrelation(fit = results2$model)

# checking auto-correlation estimates
ar_check2$ar_summary
```

If the AR term correctly captured residual auto-correlation, we expect the average lag-1 auto-correlation to be lower in the corrected residuals than in the raw residuals.

```{r print-ar2}
head(x = ar_check2$rho1_by_series_raw)
head(x = ar_check2$rho1_by_series_corrected)

# average lag-1 autocorrelation across series
mean(x = ar_check2$rho1_by_series_raw$rho1, na.rm = TRUE)
mean(x = ar_check2$rho1_by_series_corrected$rho1, na.rm = TRUE)
```

We then visualise the identified clusters, showing that this model produces much more conservative estimates (because of the larger uncertainty about group-level estimates.)

```{r print-clusters2}
# displaying the identified clusters
print(results2)
```

```{r clusters-plot2, fig.width = 9, fig.height = 6}
# plotting the data, model's predictions, and clusters
plot(results2)
```

We visualise both the auto-correlation (ACF) and partial auto-correlation (pACF) of the raw and AR-corrected (whitened) residuals.

```{r clusters-plots-combined2, fig.width = 12, fig.height = 12}
(ar_check2$plots$acf_raw + ar_check2$plots$pacf_raw) /
    (ar_check2$plots$acf_corrected + ar_check2$plots$pacf_corrected)
```

## Model comparison

Model comparison suggests that the second model has much better out-of-sample predictive accuracy (lower WAIC).

```{r model-comparison}
# model comparison using WAIC
brms::waic(results1$model, results2$model)
```

## Lowering auto-correlation with varying smooth

Another way of reducing residual auto-correlation is to improve the model fit to capture finer structure present in the data. Below we show that auto-correlation can be considerably lowered by including per-participant varying smoothers.

```{r clusters3, results = "hide"}
# fitting the BGAMM to identify clusters
results3 <- testing_through_time(
    # EEG data
    data = eeg_data,
    # participant column
    participant_id = "subject",
    # EEG column
    outcome_id = "voltage",
    # no predictor
    predictor_id = NA,
    # basis dimension
    kvalue = 20,
    # we recommend fitting the GAMM on summary statistics (mean and SD)
    multilevel = "summary",
    # no by-participant varying smooth (for speed purposes)
    varying_smooth = TRUE,
    # threshold on posterior odds
    threshold = 10,
    # number of MCMCs
    chains = 4,
    # number of parallel cores
    cores = 4,
    # number of warmup iterations
    warmup = 1000,
    # total number of iterations (per chain)
    iter = 2000
    )
```

```{r check-autocor3, fig.width = 12, fig.height = 9}
# checking residual auto-correlation
ar_check3 <- check_residual_autocorrelation(fit = results3$model)

# residual ACF and pACF (for some participants)
ar_check3$plots$acf_raw + ar_check3$plots$pacf_raw
```

```{r print-clusters3}
# displaying the identified clusters
print(results3)
```

```{r clusters-plot3, fig.width = 9, fig.height = 6}
# plotting the data, model's predictions, and clusters
plot(results3)
```

## Model comparison

Model comparison suggests that this last model has slightly better out-of-sample predictive accuracy (lower WAIC).

```{r model-comparison2}
# model comparison using WAIC
brms::waic(results1$model, results2$model, results3$model)
```
