---
title: "Modelling cross-temporal decoding generalisation matrices (2D temporal)"
author: "Ladislas Nalborczyk"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Modelling cross-temporal decoding generalisation matrices (2D temporal)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    cache = TRUE,
    collapse = TRUE,
    warning = FALSE,
    message = FALSE,
    comment = "#>",
    out.width = "100%",
    dev = "png",
    dpi = 300
    )
```

## Importing and visualising the data

Below we import and visualise the cross-temporal decoding generalisation data discussed in Nalborczyk & Bürkner ([2025, Appendix A](https://www.biorxiv.org/content/10.1101/2025.08.29.672336v1)).

```{r data-import, fig.width = 7, fig.height = 7}
library(neurogam)
library(ggplot2)
library(scico)
library(dplyr)

# retrieving some EEG data
data(timegen_data)
head(timegen_data)

# plotting the cross-temporal generalisation matrix
timegen_data |>
    summarise(auc = mean(auc), .by = c(train_time, test_time) ) |>
    ggplot(aes(x = train_time, y = test_time, fill = auc) ) +
    geom_tile() +
    geom_vline(xintercept = 0, linetype = 2, linewidth = 0.5) +
    geom_hline(yintercept = 0, linetype = 2, linewidth = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = 2, linewidth = 0.5) +
    scale_x_continuous(expand = c(0, 0) ) +
    scale_y_continuous(expand = c(0, 0) ) +
    coord_fixed() +
    theme_bw() +
    scico::scale_fill_scico(palette = "vik", midpoint = 0.5, name = "AUC") +
    labs(x = "Testing time (s)", y = "Training time (s)")
```

## Fitting the model

We fit a 2D BGAMM (with a varying intercept, but no varying smooth) to identify clusters with above-chance decoding accuracy. Such models are much slower than 1D temporal models, we recommend running less chains and using within-chain parallelisation (as shown below with `threads = threading(threads = 4)`) We also recommend to start with simple models (e.g., low `kvalue`, no varying effect) to be able to diagnose potential issues rapidly. For more details on such models, see for instance [Pedersen et al. (2019)](https://peerj.com/articles/6876/).

```{r model, eval = FALSE}
library(brms)

# fitting a BGAMM with two temporal dimensions
timegen_gam <- brm(
    formula = auc ~ t2(train_time, test_time, bs = "tp", k = 25) + (1 | participant),
    data = timegen_data,
    warmup = 1000,
    iter = 3000,
    chains = 2,
    cores = 2,
    backend = "cmdstanr",
    threads = threading(threads = 4),
    stan_model_args = list(stanc_options = list("O1") )
    )
```

## Identifying clusters

The previous model is equivalent to the following `testing_through_time()` call. This model takes around 15 hours on 8 parallel apple M4 cores.

```{r clusters, results = "hide"}
# we compute the smallest effect size of interest (SESOI) as the baseline
# (i.e., before stimulus onset) upper 95% quantile of AUC values
baseline_upper_quantile <- timegen_data %>%
    filter(train_time < 0 & test_time < 0) %>%
    summarise(quantile(x = auc, probs = 0.95) ) %>%
    pull()

# fitting the BGAMM to identify clusters
results <- testing_through_time(
    # cross-temporal generalisation data
    data = timegen_data,
    # participant column
    participant_id = "participant",
    # AUC column
    outcome_id = "auc",
    # testing against chance level
    predictor_id = NA,
    # defining training and testing times
    time_id = c("train_time", "test_time"),
    # basis dimension
    kvalue = 25,
    # only includes a per-participant varying intercept (no varying smooth)
    multilevel = "summary",
    use_se = FALSE,
    varying_smooth = FALSE,
    # defining chance level for AUC values
    # chance_level = 0.5,
    # defining chance level + SESOI (upper AUC values during baseline)
    chance_level = baseline_upper_quantile,
    # threshold on posterior odds
    threshold = 10,
    # testing whether AUC is above chance
    threshold_type = "above",
    # number of warmup iterations
    warmup = 1000,
    # total number of iterations (per chain)
    iter = 3000,
    # number of MCMCs
    chains = 2,
    # number of parallel cores
    cores = 2,
    # using with-chain parallelisation (needs 2x4 parallel cores)
    threads = brms::threading(threads = 4),
    # increasing max_treedepth
    stan_control = list(max_treedepth = 15)
    )
```

The results can be summarised and visualised using the dedicated `print()`, `summary()`, and `plot()` methods.

```{r clusters-show, fig.width = 12, fig.height = 6}
# print a summary table ("value" represents the posterior odds)
print(results)

# for available colour palettes, see scico::scico_palette_show()
plot(
    x = results,
    palette = "vik", midpoint = 0.5,
    axes_labels = c("Training time (s)", "Testing time (s)")
    )
```
